Comparing and combining network-guided biomarker discovery to discover genetic susceptibility mechanisms to breast cancer on the GENESIS study

### 2021-01-14 manuscript/plos/review_2.tex
revisions 2: added manuscript and some changes

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{Biological networks and GWAS: comparing and combining network methods to understand the genetics of familial breast cancer susceptibility in the GENESIS study} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Héctor Climente-González\textsuperscript{1,2,3,4*}, 
Christine Lonjou\textsuperscript{1,2,3}, 
Fabienne Lesueur\textsuperscript{1,2,3},
Dominique Stoppa-Lyonnet\textsuperscript{5,6,7\textcurrency}, 
Nadine Andrieu\textsuperscript{1,2,3}, 
Chloé-Agathe Azencott\textsuperscript{3,1,2},
GENESIS study group\textsuperscript{\textpilcrow}
\\
\bigskip
\textbf{1} Institut Curie, PSL Research University, F-75005 Paris, France;\\
\textbf{2} INSERM, U900, F-75005 Paris, France;\\
\textbf{3} MINES ParisTech, PSL Research University, CBIO-Centre for Computational Biology, F-75006 Paris, France;\\
\textbf{4} RIKEN Center for Advanced Intelligence Project (AIP), Tokyo, Japan; \\
\textbf{5} Service de Génétique, Institut Curie, F-75005 Paris, France;\\
\textbf{6} INSERM, U830, F-75005 Paris, France;\\
\textbf{7} Université Paris Descartes.\\
\bigskip

@@

In this work, we studied the relevance of network-based approaches to genetics by applying these six network methods to GWAS data. They use different interpretations of the omnigenic model and provide a representative view of the field. We worked on the GENESIS dataset \cite{sinilnikova_genesis:_2016}, a study on familial breast cancer conducted in the French population. After a classical GWAS approach, we used these network methods to identify additional breast cancer susceptibility genes. Lastly, we compared the solutions obtained by the different methods and aggregated them to obtain consensus solutions of predisposition to familial breast cancer that addressed their shortcomings.

@@

To measure the association between genotype and susceptibility to breast cancer, we performed a per-SNP 1~d.f.~$\chi^2$ allelic test using PLINK v1.90 \cite{chang_second-generation_2015}. To obtain significant SNPs, we performed a Bonferroni correction to keep the family-wise error rate below 5\%. The threshold used was $\frac {0.05} {197083} = 2.54 \times 10^{-7}$.

Then, we used VEGAS2 \cite{mishra_vegas2:_2015} to compute the gene-level association score from the P\=/values of the SNPs mapped to them. More specifically, we mapped SNPs to genes through their genomic coordinates: all SNPs located within the boundaries of a gene, \textpm 50 kb, were mapped to that gene. We computed VEGAS2 scores for each gene using only the 10\% of SNPs with the lowest P\=/values among all those mapped to it. We used the 62\,193 genes described in GENCODE 31 \cite{frankish_gencode_2019}, although only 54\,612 mapped to at least one SNP. Out of those, we focused exclusively on the 32\,767 that had a gene symbol. Out of the 197\,083 SNPs remaining after quality control, 164\,037 mapped to at least one of these genes. We also performed a Bonferroni correction to obtain significant genes; in this case, the threshold of significance was $\frac {0.05} {32767} = 1.53 \times 10^{-6}$.

@@

We searched for pathways enriched in the gene solutions produced by the above methods. We conducted a hypergeometric test on pathways from Reactome \cite{Jassal2019} using the function \texttt{enrichPathway} from the \texttt{ReactomePA} R package \cite{Yu2016}. The universe of genes included any gene that we could map to a SNP in the iCOGS array (Section~\ref{methods:conventional}). We adjusted the P-values for multiple testing as in Benjamini and Hochberg \cite{Benjamini1995} (BH): pathways with a BH adjusted P\=/value \textless~0.05 were deemed significant.

@@

Rewiring the PPIN while preserving the number of edges of each gene allowed to study the impact of the topology on the output of network methods. Indeed, the edges lose their biological meaning while the topology of the network is conserved. We produced 100 such rewirings by randomly swapping edges in the PPIN. We still scored the genes as described in Section~\ref{methods:methods}. We only applied only four methods on the rewirings: heinz, dmGWAS, LEAN, and SigMod. We excluded HotNet2 and SConES since they took notably longer to run.

@@

Lastly, a pathway enrichment analysis (Section~\ref{methods:pathway_enrichment}) also showed similarities and differences between the methods' solutions. It linked different parts of SigMod's solution to four processes (\nameref{stab:sigmod_pwy}): protein translation (including mitochondrial), mRNA splicing, protein misfolding, and keratinization (BH adjusted P\=/values~$<$~0.03). Interestingly, the dmGWAS solution (\nameref{stab:dmgwas_pwy}) was also related to protein misfolding (\emph{attenuation phase}, BH adjusted P\=/value~=~0.01). However, it additionally included submodules of proteins related to mitosis, DNA damage, and regulation of TP53 (BH adjusted P\=/values~$<$~0.05), which match previously known mechanisms of breast cancer susceptibility \cite{nielsen_hereditary_2016}. As with SigMod, the genes in HotNet2's solution (\nameref{stab:hotnet2_pwy}) were involved in mitochondrial translation (BH adjusted P\=/value~=~1.87 \texttimes{} 10\textsuperscript{-4}), but also in glycogen metabolism and transcription of nuclear receptors (BH adjusted P\=/value~$<$~0.04).

@@

\begin{figure}[!ht]
  \centering
  \includegraphics[width=.9\linewidth]{./figures/figure_3.pdf}
  \caption{ \textbf{Consensus solution on GENESIS (Section~\ref{methods:methods}).} \textbf{(A)}~Manhattan plot of genes; in black, the ones in the consensus solution. The red line indicates the Bonferroni threshold (1.53~\texttimes{}~10\textsuperscript{-6} for genes). \textbf{(B)}~Consensus network. Each gene is represented by a pie chart, which shows the methods that selected it. We enlarged the two most central genes (\emph{COPS5} and \emph{OFD1}), the known breast cancer susceptibility genes, and the BCAC-significant genes (Section~\ref{methods:bcac}). \textbf{(C)}~The nodes are in the same disposition as in panel B, but we indicated every gene name. We colored in pink the names of known breast cancer susceptibility genes and BCAC-significant genes.}
  \label{fig:consensus}
\end{figure}

@@

\begin{figure}[!ht]
  \centering
  \includegraphics[width=.8\linewidth]{./figures/figure_4.pdf}
  \caption{\textbf{Comparison of network methods on GENESIS.} Each method was run 5 times on a random subset containing 80\% of the samples and tested on the remaining samples (Section~\ref{methods:benchmark}). As LEAN did not select any gene, we excluded it from panels A and B. \textbf{(A)}~Number of SNPs selected by each method and number of SNPs in the active set (i.e., the number of SNPs selected by the classifier, Section~\ref{methods:classifier}). Points are the average over the 5 runs; the error bars represent the standard error of the mean. A grey diagonal line with slope 1 is added for comparison, indicating the upper bound of the active set. For reference, the active set of the classifier using all the SNPs as input included, on average, 154\,117.4 SNPs. \textbf{(B)}~Pairwise Pearson correlations of the solutions produced by different methods. A Pearson correlation of 1 means the two solutions are the same. \textbf{(C)}~Runtime of the evaluated methods, by type of network used (PPIN or SNP). For gene-based methods, inverted triangles represent the runtime of the algorithm alone, and circles the total time, which includes the algorithm themselves and the additional 119\,980 seconds (1 day and 9.33 hours) that VEGAS2 took on average to compute the gene scores from SNP summary statistics. \textbf{(D)}~True positive rate and true negative rate of the methods, obtained using different parameter combinations (Section~\ref{results:parameters}). We used as true positives BCAC-significant SNPs (for SConES and $\chi^2$ + Bonferroni) and genes (for the remaining methods, Section~\ref{methods:bcac}). We used the whole dataset in this panel.}
  \label{fig:benchmark}
  \end{figure}

@@

\begin{figure}[!ht]
  \centering
  \includegraphics[width=.8\linewidth]{./figures/figure_2.pdf}
  \caption{\textbf{Drawbacks encountered when using network methods.} \textbf{(A)}~DmGWAS solution, with the genes colored according to the -log$_{10}$ of their P\=/value. \textbf{(B)}~Number of times a gene was selected by either dmGWAS, heinz, LEAN, or SigMod in 100 rewirings of the PPIN (Section~\ref{methods:pathway_enrichment}) and its centrality degree. \textbf{(C)}~Centrality degree and -log\textsubscript{10} of the VEGAS2 P\=/value in BCAC for each of the nodes in the PPIN. We highlighted the genes selected by each method and the ones selected by more than one (``Consensus''). We labeled the three most central genes that were picked by any method. \textbf{(D)}~Overlap between the solutions of SConES GS, GM, or GI. Barplots are colored based on whether the SNPs map to a gene or not (Section~\ref{methods:snp2gene}).}
  \label{fig:issues}
  \end{figure}

@@

By using a SNP subnetwork, SConES analyzed each SNP in its functional context. Therefore, it could select SNPs located in genes not included in the PPIN and in non-coding regions. We compared the solution of SConES in the GI network (using PPIN information), to the one using only positional information (GS network) and to the one using positional and gene annotations (GM network). Importantly, SConES produced similar results on the GS and GM networks (\nameref{sfig:pearson_methods}). While the solutions on those two considerably overlap with SConES GI's, they contained additional gene-coding segments (Fig~\ref{fig:issues}C). In fact, both SConES GS and GM selected chromosome regions related to breast cancer, like 3p24 (\emph{SLC4A7}/\emph{NEK10} \cite{ahmed_newly_2009}), 5p12 (\emph{FGF10}, \emph{MRPS30} \cite{quigley_5p12_2014}), 10q26 (\emph{FGFR2}), and 16q12 (\emph{TOX3}). In addition to those, SConES GS selected region 8q24, also linked to breast cancer (\emph{POU5F1B} \cite{breyer_expressed_2014}).

@@

To overcome these limitations while exploiting the each method's strengths, we proposed combining them into a consensus solution. We used the straightforward strategy of including any node that was recovered by at least two methods. We thus proposed two solutions (Sections~\ref{results:consensus} and \ref{results:stable-consensus}): a consensus solution, which addressed the heterogeneity of the solutions, and a stable consensus solution, which addresseded the instability of the methods. They both included the majority of the strongly associated smaller solutions and captured genes and broader mechanisms related to cancer, thus synthesizing the mechanisms altered in breast cancer cases. Thanks to their smaller size and network structure, they provided compelling hypotheses on genes like \emph{COPS5} and \emph{CUL3}, which lack genome-wide association with the disease but are related to cancer at the expression level and consistently interact with high scoring genes. Notably, while the consensus approach was as unstable as the individual network-guided methods, the stable consensus network retained the ability to provide compelling hypotheses and had better stability. This supported that redundant but equivalent mechanisms might cause instability and supported the conclusions obtained on the individual solutions.

@@

% Include only the SI item label in the paragraph heading. Use the \nameref{label} command to cite SI items in the text.
\paragraph*{S1 Table.}
\label{stab:snp_solutions}
\textbf{Summary statistics on the results of SConES on the three SNP networks (Section~\ref{methods:networks}).} The first row within each block contains the summary statistics on the whole network.

\paragraph*{S2 Table.}
\label{stab:sigmod_pwy}
\textbf{Pathway enrichment analyses of the genes in SigMod's solution.}

\paragraph*{S3 Table.}
\label{stab:dmgwas_pwy}
\textbf{Pathway enrichment analyses of the genes in dmGWAS' solution.}

\paragraph*{S4 Table.}
\label{stab:hotnet2_pwy}
\textbf{Pathway enrichment analyses of the genes in HotNet2's solution.}

\paragraph*{S5 Table.}
\label{stab:consensus_pwy}
\textbf{Pathway enrichment analyses of the genes in the consensus' solution.}

@@

\paragraph*{S5 Fig.}
\label{sfig:pearson_methods}
\textbf{Pearson correlation between the different solutions.} \textbf{(A)}~Correlation between selected SNPs. \textbf{(B)}~Correlation between selected genes. In general, the solutions display a very low overlap.
%\includegraphics[width=.9\linewidth]{./figures/sfigure_5.pdf}

\paragraph*{S6 Fig.}
\label{sfig:rewirings}
\textbf{Number of times a gene was selected by either dmGWAS, heinz, LEAN, or SigMod in 100 rewirings of the PPIN (Section~\ref{methods:rewiring}) and its centrality degree.} This figure is equivalent to Fig~\ref{fig:issues}B, split by method. 
%\includegraphics[width=.9\linewidth]{./figures/sfigure_6.pdf

@@

\paragraph*{S9 Fig.}
\label{sfig:biotypes_excluded}
\textbf{Biotypes of genes from the annotation that are not present in the HINT PPIN.}
%\includegraphics[width=.9\linewidth]{./figures/sfigure_9.pdf}

\paragraph*{S10 Fig.}
\label{sfig:lc_ht_comparison}
\textbf{Comparison of the benchmark on high-throughput (HT) interactions to the benchmark on both high-throughput and literature curated interactions (HT+LC).} Grey lines represent no change in the statistic between the benchmarks (1 for ratios mean(HT) / mean(HT + LC), 0 for differences mean(HT) - mean(HT + LC)). \textbf{(A)}~Ratios of the selected features between both benchmarks and of the active set (Section~\ref{methods:classifier}). \textbf{(B)}~Shifts in sensitivity and specificity. \textbf{(C)}~Shift in Pearson correlation between benchmarks. \textbf{(D)}~Ratio between the runtimes of the benchmarks. For gene-based methods, inverted triangles represent the ratio of runtimes of the algorithms themselves, and circles the total time, which includes the algorithm themselves and the additional 119\,980 seconds (1 day and 9.33 hours) that VEGAS2 took on average to compute the gene scores from SNP summary statistics. In general, adding additional interactions slightly improved the stability of the solution. However, it increased the solution size and the required runtime, and had mixed effects on the sensitivity and specificity.
%\includegraphics[width=.85\linewidth]{./figures/sfigure_10.pdf}

@@

We thank all the GENESIS collaborating cancer clinics clinics (Clinique Sainte Catherine, Avignon: H. Dreyfus; Hôpital Saint Jacques, Besançon: M-A. Collonge-Rame; Institut Bergonié, Bordeaux: M.Longy, A. Floquet, E. Barouk-Simonet; CHU, Brest: S. Audebert; Centre François Baclesse, Caen: P. Berthet; Hôpital Dieu, Chambéry: S. Fert-Ferrer; Centre Jean Perrin, Clermont-Ferrand: Y-J. Bignon; Hôpital Pasteur, Colmar: J-M. Limacher; Hôpital d’Enfants CHU – Centre Georges François Leclerc, Dijon: L. Faivre-Olivier; CHU, Fort de France: O. Bera; CHU Albert Michallon, Grenoble: D. Leroux; Hôpital Flaubert, Le Havre: V. Layet; Centre Oscar Lambret, Lille: P. Vennin, C. Adenis; Hôpital Jeanne de Flandre, Lille: S. Lejeune-Dumoulin, S. Manouvier-Hanu; CHRU Dupuytren, Limoges: L. Venat-Bouvet; Centre Léon Bérard, Lyon: C. Lasset, V. Bonadona; Hôpital Edouard Herriot, Lyon: S. Giraud; Institut Paoli-Calmettes, Marseille: F. Eisinger, L. Huiart; Centre Val d’Aurelle – Paul Lamarque, Montpellier: I. Coupier; CHU Arnaud de Villeneuve, Montpellier: I. Coupier, P. Pujol; Centre René Gauducheau, Nantes: C. Delnatte; Centre Catherine de Sienne, Nantes: A. Lortholary; Centre Antoine Lacassagne, Nice: M. Frénay, V. Mari; Hôpital Caremeau, Nîmes: J. Chiesa; Réseau Oncogénétique Poitou Charente, Niort: P. Gesta; Institut Curie, Paris: D. Stoppa-Lyonnet, M. Gauthier-Villars, B. Buecher, A. de Pauw, C. Abadie, M. Belotti; Hôpital Saint-Louis, Paris: O. Cohen-Haguenauer; Centre Viggo-Petersen, Paris: F. Cornélis; Hôpital Tenon, Paris: A. Fajac; GH Pitié Salpétrière et Hôpital Beaujon, Paris: C. Colas, F. Soubrier, P. Hammel, A. Fajac; Institut Jean Godinot, Reims: C. Penet, T. D. Nguyen; Polyclinique Courlancy, Reims: L. Demange*, C. Penet; Centre Eugène Marquis, Rennes: C. Dugast*; Centre Henri Becquerel, Rouen: A. Chevrier, T. Frebourg, J. Tinat, I. Tennevet, A. Rossi; Hôpital René Huguenin/Institut Curie, Saint Cloud: C. Noguès, L. Demange*, E. Mouret-Fourme; CHU, Saint-Etienne: F. Prieur; Centre Paul Strauss, Strasbourg: J-P. Fricker, H. Schuster; Hôpital Civil, Strasbourg: O. Caron, C. Maugard; Institut Claudius Regaud, Toulouse: L. Gladieff, V. Feillel; Hôpital Bretonneau, Tours: I. Mortemousque; Centre Alexis Vautrin, Vandoeuvre-les-Nancy: E. Luporsi; Hôpital de Bravois, Vandoeuvre-les-Nancy: P. Jonveaux; Gustave Roussy, Villejuif: A. Chompret*, O. Caron).
*Deceased prematurely

>>>>

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{Biological networks and GWAS: comparing and combining network methods to understand the genetics of familial breast cancer susceptibility in the GENESIS study} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Héctor Climente-González\textsuperscript{1,2,3,4*}, 
Christine Lonjou\textsuperscript{1,2,3}, 
Fabienne Lesueur\textsuperscript{1,2,3},
GENESIS study group\textsuperscript{\textpilcrow},
Dominique Stoppa-Lyonnet\textsuperscript{5,6,7\textcurrency}, 
Nadine Andrieu\textsuperscript{1,2,3}, 
Chloé-Agathe Azencott\textsuperscript{3,1,2}
\\
\bigskip
\textbf{1} Institut Curie, PSL Research University, F-75005 Paris, France;\\
\textbf{2} INSERM, U900, F-75005 Paris, France;\\
\textbf{3} MINES ParisTech, PSL Research University, CBIO-Centre for Computational Biology, F-75006 Paris, France;\\
\textbf{4} RIKEN Center for Advanced Intelligence Project (AIP), Tokyo, Japan; \\
\textbf{5} Service de Génétique, Institut Curie, F-75005 Paris, France;\\
\textbf{6} INSERM, U830, F-75005 Paris, France;\\
\textbf{7} Université Paris Descartes.\\
\bigskip

@@

In this work, we studied the relevance of network-based approaches to genetics by applying these six network methods to GWAS data. They use different interpretations of the omnigenic model and provide a representative view of the field. We worked on the GENESIS dataset \cite{sinilnikova_genesis:_2016}, a study on familial breast cancer conducted in the French population. After a classical GWAS approach, we used these network methods to identify additional breast cancer susceptibility genes. Lastly, we compared the solutions obtained by the different methods and computed their intersection to obtain consensus solutions of predisposition to familial breast cancer that addressed their shortcomings.

@@

To measure the association between genotype and susceptibility to breast cancer, we performed a per-SNP 1~d.f.~$\chi^2$ allelic test using PLINK v1.90 \cite{chang_second-generation_2015}. To obtain significant SNPs, we performed a Bonferroni correction to keep the family-wise error rate below 5\%. The threshold used was $\frac {0.05} {197083} = 2.54 \times 10^{-7}$. The summary statistics of the analyzed SNPs are available in \nameref{stab:snp_summary}.

Then, we used VEGAS2 \cite{mishra_vegas2:_2015} to compute the gene-level association score from the P\=/values of the SNPs mapped to them. More specifically, we mapped SNPs to genes through their genomic coordinates: all SNPs located within the boundaries of a gene, \textpm 50 kb, were mapped to that gene. We computed VEGAS2 scores for each gene using only the 10\% of SNPs with the lowest P\=/values among all those mapped to it. We used the 62\,193 genes described in GENCODE 31 \cite{frankish_gencode_2019}, although only 54\,612 mapped to at least one SNP. Out of those, we focused exclusively on the 32\,767 that had a gene symbol. Out of the 197\,083 SNPs remaining after quality control, 164\,037 mapped to at least one of these genes. We also performed a Bonferroni correction to obtain significant genes; in this case, the threshold of significance was $\frac {0.05} {32767} = 1.53 \times 10^{-6}$. The summary statistics of the analyzed genes are available in \nameref{stab:gene_summary}.

@@

We searched for pathways enriched in the solution of each network method. We conducted a hypergeometric test on pathways from Reactome \cite{Jassal2019} using the function \texttt{enrichPathway} from the \texttt{ReactomePA} R package \cite{Yu2016}. The universe of genes included any gene that we could map to a SNP in the iCOGS array (Section~\ref{methods:conventional}). We adjusted the P-values for multiple testing as in Benjamini and Hochberg \cite{Benjamini1995} (BH): pathways with a BH adjusted P\=/value \textless~0.05 were deemed significant.

@@

Rewiring the PPIN while preserving the number of edges of each gene allowed to study the impact of the topology on the output of network methods. Indeed, the edges lose their biological meaning while the topology of the network is conserved. We produced 100 such rewirings by randomly swapping edges in the PPIN. We still scored the genes as described in Section~\ref{methods:methods}. We applied only four methods on the rewirings: heinz, dmGWAS, LEAN, and SigMod. We excluded HotNet2 and SConES since they took notably longer to run. As on the real PPIN, LEAN did not produce significant results in the rewirings.

@@

Lastly, pathway enrichment analyses on each method's solution (Section~\ref{methods:pathway_enrichment}) also revealed similarities and differences between them. It linked different parts of SigMod's solution to four processes (\nameref{stab:sigmod_pwy}): protein translation (including mitochondrial), mRNA splicing, protein misfolding, and keratinization (BH adjusted P\=/values~$<$~0.03). Interestingly, the dmGWAS solution (\nameref{stab:dmgwas_pwy}) was also related to protein misfolding (\emph{attenuation phase}, BH adjusted P\=/value~=~0.01). However, it additionally included proteins related to mitosis, DNA damage, and regulation of TP53 (BH adjusted P\=/values~$<$~0.05), which match previously known mechanisms of breast cancer susceptibility \cite{nielsen_hereditary_2016}. As with SigMod, the genes in HotNet2's solution (\nameref{stab:hotnet2_pwy}) were involved in mitochondrial translation (BH adjusted P\=/value~=~1.87 \texttimes{} 10\textsuperscript{-4}), but also in glycogen metabolism and transcription of nuclear receptors (BH adjusted P\=/value~$<$~0.04).

@@

\begin{figure}[!ht]
  \centering
  \includegraphics[width=.9\linewidth]{./figures/figure_2.pdf}
  \caption{ \textbf{Consensus solution on GENESIS (Section~\ref{methods:methods}).} \textbf{(A)}~Manhattan plot of genes; in black, the ones in the consensus solution. The red line indicates the Bonferroni threshold (1.53~\texttimes{}~10\textsuperscript{-6} for genes). \textbf{(B)}~Consensus network. Each gene is represented by a pie chart, which shows the methods that selected it. We enlarged the two most central genes (\emph{COPS5} and \emph{OFD1}), the known breast cancer susceptibility genes, and the BCAC-significant genes (Section~\ref{methods:bcac}). \textbf{(C)}~The nodes are in the same disposition as in panel B, but we indicated every gene name. We colored in pink the names of known breast cancer susceptibility genes and BCAC-significant genes.}
  \label{fig:consensus}
\end{figure}

@@

\begin{figure}[!ht]
  \centering
  \includegraphics[width=.8\linewidth]{./figures/figure_3.pdf}
  \caption{\textbf{Comparison of network methods on GENESIS.} Each method was run 5 times on a random subset containing 80\% of the samples and tested on the remaining samples (Section~\ref{methods:benchmark}). As LEAN did not select any gene, we excluded it from panels A and B. \textbf{(A)}~Number of SNPs selected by each method and number of SNPs in the active set (i.e., the number of SNPs selected by the classifier, Section~\ref{methods:classifier}). Points are the average over the 5 runs; the error bars represent the standard error of the mean. A grey diagonal line with slope 1 is added for comparison, indicating the upper bound of the active set. For reference, the active set of the classifier using all the SNPs as input included, on average, 154\,117.4 SNPs. \textbf{(B)}~Pairwise Pearson correlations of the solutions produced by different methods. A Pearson correlation of 1 means the two solutions are the same. \textbf{(C)}~Runtime of the evaluated methods, by type of network used (PPIN or SNP). For gene-based methods, inverted triangles represent the runtime of the algorithm alone, and circles the total time, which includes the algorithm themselves and the additional 119\,980 seconds (1 day and 9.33 hours) that VEGAS2 took on average to compute the gene scores from SNP summary statistics. \textbf{(D)}~True positive rate and false positive rate of the methods, obtained using different parameter combinations (Section~\ref{results:parameters}). We used as true positives BCAC-significant SNPs (for SConES and $\chi^2$ + Bonferroni) and genes (for the remaining methods, Section~\ref{methods:bcac}). We used the whole dataset in this panel.}
  \label{fig:benchmark}
  \end{figure}

@@

\begin{figure}[!ht]
  \centering
  \includegraphics[width=.8\linewidth]{./figures/figure_4.pdf}
  \caption{\textbf{Drawbacks encountered when using network methods.} \textbf{(A)}~DmGWAS solution, with the genes colored according to the -log$_{10}$ of their P\=/value. \textbf{(B)}~Number of times a gene was selected by either dmGWAS, heinz, LEAN, or SigMod in 100 rewirings of the PPIN (Section~\ref{methods:pathway_enrichment}) and its centrality degree. \textbf{(C)}~Centrality degree and -log\textsubscript{10} of the VEGAS2 P\=/value in BCAC for each of the nodes in the PPIN. We highlighted the genes selected by each method and the ones selected by more than one (``Consensus''). We labeled the three most central genes that were picked by any method. \textbf{(D)}~Overlap between the solutions of SConES GS, GM, or GI. Barplots are colored based on whether the SNPs map to a gene or not (Section~\ref{methods:snp2gene}).}
  \label{fig:issues}
  \end{figure}

@@

By using a SNP subnetwork, SConES analyzed each SNP in its functional context. Therefore, it could select SNPs located in genes not included in the PPIN and in non-coding regions. We compared the solution of SConES in the GI network (using PPIN information), to the one using only positional information (GS network) and to the one using positional and gene annotations (GM network). Importantly, SConES produced similar results on the GS and GM networks (\nameref{sfig:pearson_methods}). While the solutions on those two considerably overlap with SConES GI's, they contained additional gene-coding segments (Fig~\ref{fig:issues}D). In fact, both SConES GS and GM selected chromosome regions related to breast cancer, like 3p24 (\emph{SLC4A7}/\emph{NEK10} \cite{ahmed_newly_2009}), 5p12 (\emph{FGF10}, \emph{MRPS30} \cite{quigley_5p12_2014}), 10q26 (\emph{FGFR2}), and 16q12 (\emph{TOX3}). In addition to those, SConES GS selected region 8q24, also linked to breast cancer (\emph{POU5F1B} \cite{breyer_expressed_2014}).

@@

To overcome these limitations while exploiting each method's strengths, we proposed combining them into a consensus solution. We used the straightforward strategy of including any node that was recovered by at least two methods. We thus proposed two solutions (Sections~\ref{results:consensus} and \ref{results:stable-consensus}): a consensus solution, which addressed the heterogeneity of the solutions, and a stable consensus solution, which addresseded the instability of the methods. They both included the majority of the strongly associated smaller solutions and captured genes and broader mechanisms related to cancer, thus synthesizing the mechanisms altered in breast cancer cases. Thanks to their smaller size and network structure, they provided compelling hypotheses on genes like \emph{COPS5} and \emph{CUL3}, which lack genome-wide association with the disease but are related to cancer at the expression level and consistently interact with high scoring genes. Notably, while the consensus approach was as unstable as the individual network-guided methods, the stable consensus network retained the ability to provide compelling hypotheses and had better stability. This supported that redundant but equivalent mechanisms might cause instability and supported the conclusions obtained on the individual solutions.

@@

% Include only the SI item label in the paragraph heading. Use the \nameref{label} command to cite SI items in the text.
\paragraph*{S1 Table.}
\label{stab:snp_summary}
\textbf{SNP summary statistics on GENESIS.}

\paragraph*{S2 Table.}
\label{stab:gene_summary}
\textbf{Gene summary statistics on GENESIS.}

\paragraph*{S3 Table.}
\label{stab:snp_solutions}
\textbf{Summary statistics on the results of SConES on the three SNP networks (Section~\ref{methods:networks}).} The first row within each block contains the summary statistics on the whole network.

\paragraph*{S4 Table.}
\label{stab:sigmod_pwy}
\textbf{Pathway enrichment analyses of the genes in SigMod's solution.}

\paragraph*{S5 Table.}
\label{stab:dmgwas_pwy}
\textbf{Pathway enrichment analyses of the genes in dmGWAS' solution.}

\paragraph*{S6 Table.}
\label{stab:hotnet2_pwy}
\textbf{Pathway enrichment analyses of the genes in HotNet2's solution.}

\paragraph*{S7 Table.}
\label{stab:consensus_pwy}
\textbf{Pathway enrichment analyses of the genes in the consensus' solution.}

@@

\paragraph*{S5 Fig.}
\label{sfig:rewirings}
\textbf{Number of times a gene was selected by either dmGWAS, heinz, or SigMod in 100 rewirings of the PPIN (Section~\ref{methods:rewiring}) and its centrality degree.} This figure is equivalent to Fig~\ref{fig:issues}B, split by method. 
%\includegraphics[width=.9\linewidth]{./figures/sfigure_5.pdf

\paragraph*{S6 Fig.}
\label{sfig:pearson_methods}
\textbf{Pearson correlation between the different solutions.} \textbf{(A)}~Correlation between selected SNPs. \textbf{(B)}~Correlation between selected genes. In general, the solutions display a very low overlap.
%\includegraphics[width=.9\linewidth]{./figures/sfigure_6.pdf}

@@

\paragraph*{S9 Fig.}
\label{sfig:lc_ht_comparison}
\textbf{Comparison of the benchmark on high-throughput (HT) interactions to the benchmark on both high-throughput and literature curated interactions (HT+LC).} Grey lines represent no change in the statistic between the benchmarks (1 for ratios mean(HT) / mean(HT + LC), 0 for differences mean(HT) - mean(HT + LC)). \textbf{(A)}~Ratios of the selected features between both benchmarks and of the active set (Section~\ref{methods:classifier}). \textbf{(B)}~Shifts in sensitivity and specificity. \textbf{(C)}~Shift in Pearson correlation between benchmarks. \textbf{(D)}~Ratio between the runtimes of the benchmarks. For gene-based methods, inverted triangles represent the ratio of runtimes of the algorithms themselves, and circles the total time, which includes the algorithm themselves and the additional 119\,980 seconds (1 day and 9.33 hours) that VEGAS2 took on average to compute the gene scores from SNP summary statistics. In general, adding additional interactions slightly improved the stability of the solution. However, it increased the solution size and the required runtime, and had mixed effects on the sensitivity and specificity.
%\includegraphics[width=.85\linewidth]{./figures/sfigure_10.pdf}

\paragraph*{S10 Fig.}
\label{sfig:biotypes_excluded}
\textbf{Biotypes of genes from the annotation that are not present in the HINT PPIN.}
%\includegraphics[width=.9\linewidth]{./figures/sfigure_9.pdf}

@@

We thank all the GENESIS collaborating cancer clinics (Clinique Sainte Catherine, Avignon: H. Dreyfus; Hôpital Saint Jacques, Besançon: M-A. Collonge-Rame; Institut Bergonié, Bordeaux: M.Longy, A. Floquet, E. Barouk-Simonet; CHU, Brest: S. Audebert; Centre François Baclesse, Caen: P. Berthet; Hôpital Dieu, Chambéry: S. Fert-Ferrer; Centre Jean Perrin, Clermont-Ferrand: Y-J. Bignon; Hôpital Pasteur, Colmar: J-M. Limacher; Hôpital d’Enfants CHU – Centre Georges François Leclerc, Dijon: L. Faivre-Olivier; CHU, Fort de France: O. Bera; CHU Albert Michallon, Grenoble: D. Leroux; Hôpital Flaubert, Le Havre: V. Layet; Centre Oscar Lambret, Lille: P. Vennin, C. Adenis; Hôpital Jeanne de Flandre, Lille: S. Lejeune-Dumoulin, S. Manouvier-Hanu; CHRU Dupuytren, Limoges: L. Venat-Bouvet; Centre Léon Bérard, Lyon: C. Lasset, V. Bonadona; Hôpital Edouard Herriot, Lyon: S. Giraud; Institut Paoli-Calmettes, Marseille: F. Eisinger, L. Huiart; Centre Val d’Aurelle – Paul Lamarque, Montpellier: I. Coupier; CHU Arnaud de Villeneuve, Montpellier: I. Coupier, P. Pujol; Centre René Gauducheau, Nantes: C. Delnatte; Centre Catherine de Sienne, Nantes: A. Lortholary; Centre Antoine Lacassagne, Nice: M. Frénay, V. Mari; Hôpital Caremeau, Nîmes: J. Chiesa; Réseau Oncogénétique Poitou Charente, Niort: P. Gesta; Institut Curie, Paris: D. Stoppa-Lyonnet, M. Gauthier-Villars, B. Buecher, A. de Pauw, C. Abadie, M. Belotti; Hôpital Saint-Louis, Paris: O. Cohen-Haguenauer; Centre Viggo-Petersen, Paris: F. Cornélis; Hôpital Tenon, Paris: A. Fajac; GH Pitié Salpétrière et Hôpital Beaujon, Paris: C. Colas, F. Soubrier, P. Hammel, A. Fajac; Institut Jean Godinot, Reims: C. Penet, T. D. Nguyen; Polyclinique Courlancy, Reims: L. Demange*, C. Penet; Centre Eugène Marquis, Rennes: C. Dugast*; Centre Henri Becquerel, Rouen: A. Chevrier, T. Frebourg, J. Tinat, I. Tennevet, A. Rossi; Hôpital René Huguenin/Institut Curie, Saint Cloud: C. Noguès, L. Demange*, E. Mouret-Fourme; CHU, Saint-Etienne: F. Prieur; Centre Paul Strauss, Strasbourg: J-P. Fricker, H. Schuster; Hôpital Civil, Strasbourg: O. Caron, C. Maugard; Institut Claudius Regaud, Toulouse: L. Gladieff, V. Feillel; Hôpital Bretonneau, Tours: I. Mortemousque; Centre Alexis Vautrin, Vandoeuvre-les-Nancy: E. Luporsi; Hôpital de Bravois, Vandoeuvre-les-Nancy: P. Jonveaux; Gustave Roussy, Villejuif: A. Chompret*, O. Caron).
*Deceased prematurely
